### 2013-01-15 edupar-13/410.tex
incorporate changes suggested in meeting

Finally, both to make up for the tests' incompleteness and to provide personalized human feedback, each student kernel is manually graded by a member of course staff.
The instructors and teaching assistants print out each group's submitted kernel and mark it with red ink, auditing the code for code style, architectural issues, race conditions, and preemptibility.
After the inking process, each group meets with the staff member who graded them for an hour-long ``kernel interview'', essentially a debriefing in which any final misconceptions the students might have can be rectified.
Staff discretionary points constitute 20\% of the kernel project's grade.

>>>>

Finally, both to make up for the tests' incompleteness and to provide personalized human feedback, each student kernel is manually graded by a member of course staff.
The instructors and teaching assistants print out each group's submitted kernel and mark it with red ink, auditing the code for style, architectural issues, race conditions, and preemptibility.
After the inking process, each group meets with the staff member who graded them for an hour-long ``kernel interview'', essentially a debriefing in which any final misconceptions the students might have can be rectified.
Staff discretionary points constitute 20\% of the kernel project's grade.

### 2013-01-15 edupar-13/conclusion.tex
incorporate changes suggested in meeting

Many people have contributed to the development of
15-410 over the past decade.
Stephen Muckle, currently at Qualcomm, led
the switch to x86 projects running in Simics.
Nathaniel Wesley Filardo,
currently a Ph.D.\ candidate at Johns Hopkins,
increased the sanity of the build infrastructure,
designed the \x{swexn()} system call, and helped critique this paper.
The current reference kernel is named ``pathos''
and was written by Michael J. Sullivan,
currently a CMU CS Ph.D.\ candidate,
and Elly Fong-Jones, currently at Google.
The SMP infrastructure code was contributed by
Ryan Pearl, currently at Mozilla.
Joshua Wise, currently at NVIDIA,
proposed and led several kernel-extension projects,
including the hypervisor project.
Over the past decade the course has benefitted
greatly from free educational Simics licenses,
donated initially by Virtutech and recently by
Wind River.
Code-size numbers were generated using David A.\ Wheeler's
``SLOCCount.''
Work on Landslide was sponsored by the U.S. Army Research Office under grant number W911NF0910273.

>>>>

Many people have contributed to the development of
15-410 over the past decade.
Stephen Muckle, currently at Qualcomm, led
the switch to x86 projects running in Simics.
Nathaniel Wesley Filardo,
currently a Ph.D.\ candidate at Johns Hopkins,
% increased the sanity of the build infrastructure,
designed the \x{swexn()} system call and helped critique this paper.
The current reference kernel is named ``pathos''
and was written by Michael J. Sullivan,
currently a CMU CS Ph.D.\ candidate,
and Elly Fong-Jones, currently at Google.
The SMP infrastructure code was contributed by
Ryan Pearl, currently at Mozilla.
Joshua Wise, currently at NVIDIA,
proposed and led several kernel-extension projects,
including the hypervisor project.
Over the past decade the course has benefitted
greatly from free educational Simics licenses
donated by Virtutech/Wind River.
Code-size numbers were generated using David A.\ Wheeler's
``SLOCCount.''
Work on Landslide was sponsored by the U.S. Army Research Office under grant number W911NF0910273.

### 2013-01-15 edupar-13/intro.tex
remove unnecessary citations
incorporate changes suggested in meeting

In this paper we describe our experience teaching concurrent programming at the operating systems level to advanced undergraduate students at Carnegie Mellon University (CMU).
In 15-410, the Operating Systems Design and Implementation class at CMU, students complete an aggressively low-level kernel project, which runs on real hardware, in which they implement threads and processes.
Presented with a kernel specification, called Pebbles~\cite{kspec}, students build a threading and synchronization library that runs in user space on top of a Pebbles-compliant kernel in a two-week project, and subsequently, in a six-week project, write their own multi-threaded kernel implementation.
During these projects students face demanding synchronization problems, including how to implement locking primitives and how to coordinate the possibly-simultaneous exiting of parent and child processes.
Finally, in another two-week project, students implement a small extension to their Pebbles kernel. This project changes every semester, and in the past has included extending their kernel serve as a hypervisor, or to support symmetric multiprocessing.

As we ask our students to write more and more difficult concurrent code in one semester, we wish to ease their burden of debugging concurrency errors, which often consumes a significant fraction of their time.
However, we also want to avoid compromising the complexity, hardware fidelity, and learning experience of the project set.
Hence, we have built Landslide~\cite{landslide}, a tool that uses systematic exploration~\cite{verisoft} to force kernel threads to take execution paths that expose concurrency bugs.
We present our experience sharing a prototype of Landslide with volunteering students who used it to find bugs in their own code, and discuss the possibility for integrating it as a main component of the 15-410 curriculum.

>>>>

In this paper we describe our experience teaching concurrent programming at the operating systems level to advanced undergraduate students at Carnegie Mellon University (CMU).
In 15-410, the Operating Systems Design and Implementation class at CMU, students complete an aggressively low-level kernel project, which runs on real hardware, in which they implement threads and processes.
Presented with a kernel specification, called Pebbles~\cite{kspec}, students build a threading and synchronization library that runs in user space on top of a Pebbles-compliant kernel in a two-week project, and subsequently, in a six-week project, write their own multi-threaded kernel implementation.
During these projects students face demanding synchronization problems, including how to implement locking primitives and how to coordinate the possibly-simultaneous exiting of parent and child processes.
Finally, in another two-week project, students implement an extension to their Pebbles kernel. This project changes every semester, and recently has involved supporting symmetric multiprocessing.
% in the past has included extending their kernel to serve as a hypervisor, or to support symmetric multiprocessing.

In comparison with other educational kernel projects,
% such as Pintos~\cite{pintos} and Xv6~\cite{xv6},
Pebbles focuses more on ground-up design than on teaching a broad feature-set.
% Pebbles allows students to design and implement their systems ``from the ground up.''
At Stanford, students build kernels following the Pintos architecture~\cite{pintos}, and at MIT, following the Xv6/JOS architecture~\cite{xv6}.
In both courses, student kernels support multiprocessor execution;
however, students are also supplied with starter code that already implements thread creation/destruction, context switching, and locking primitives.
In 15-410, we give starter code only for the machine boot sequence and for a standard library including dynamic memory allocation, leaving students to design threading and scheduling primitives from scratch.

% As we ask our students to write more and more difficult concurrent code in one semester,
As we plan to incorporate SMP into 15-410's curriculum, expanding the challenges our students face in one semester,
we wish to ease their burden of debugging concurrency errors, which often consumes a significant fraction of their time.
However, we also want to avoid compromising the complexity, hardware fidelity, and learning experience of the project set.
Hence, we have built Landslide~\cite{landslide}, a tool that uses systematic exploration~\cite{verisoft} to force kernel threads to take execution paths that expose concurrency bugs.
We present our experience sharing a prototype of Landslide with volunteering students who used it to find bugs in their own code, and discuss the possibility for integrating it as a main component of the 15-410 curriculum.

### 2013-01-15 edupar-13/landslide.tex
incorporate changes suggested in meeting

Landslide is an experimental tool for helping kernel authors find and debug concurrency errors.
It is based on the technique of systematic exploration~\cite{verisoft}, a way of exploring the state space of different possible thread interleavings in a concurrent system.
It follows in the footsteps of related tools such as
CHESS~\cite{chess} and dBug~\cite{dbug-ssv}, but is the first tool we know of to apply systematic exploration in a kernel development environment.

@@

The strategy we describe above may cause Landslide to identify decision points in unrelated parts of the kernel, such as when accessing mutexes in unrelated and/or already-trusted system calls.
We provide interface options in \x{config.landslide} for the student to view currently identified decision points and to selectively eliminate them.
Suppose a student were testing thread death and reaping, and wanted decision points to appear in \x{wait} and \x{vanish} but not if they're also associated with virtual memory operations.
They should write \x{within_function wait vanish} and \x{without_function} \x{destroy_address_space}.
The \x{within_function} directive requires that at least one of the specified functions shall be on the call stack when decision points are identified, and \x{without_function} requires the opposite.

@@

In contrast with conventional stress testing, in which the test program exercises many system calls (in parallel and/or in sequence), Landslide works best with very small test cases.
Landslide's coverage results from exploring the different scheduling possibilities that arise from a short sequence of system calls.
If Landslide were used when running a stress test, the resulting state space would be too large to explore feasibly.
Hence, when working with students, we provided a suite of four test programs, each no longer than 10 lines of code, but carefully written to bring about system call interactions known to be troublesome.

>>>>

Landslide is an experimental tool for helping kernel authors find and debug concurrency errors.
It is based on the technique of systematic exploration~\cite{verisoft}, a way of exploring the state space of different possible thread interleavings in a concurrent system.
It follows in the footsteps of related tools such as
\shortversion{
CHESS~\cite{chess},
}{
CHESS~\cite{chess}, dBug~\cite{dbug-ssv}, and DeMeter~\cite{demeter},
}
but is the first tool we know of to apply systematic exploration in a kernel development environment.

@@

The strategy we describe above may cause Landslide to identify decision points in unrelated parts of the kernel, such as when accessing mutexes in unrelated and/or already-trusted system calls.
We provide interface options in \x{config.landslide} for the student to view currently identified decision points and to selectively eliminate them.
If a student were testing thread death and reaping, they might want decision points to appear in \x{wait} and \x{vanish} but not if unrelated virtual memory operations are also in progress.
They should write \x{within_function wait vanish} and \x{without_function} \x{destroy_address_space}.
The \x{within_function} directive requires that at least one of the specified functions shall be on the call stack when decision points are identified, and \x{without_function} requires the opposite.

@@

In contrast with conventional stress testing, in which the test program exercises many system calls, Landslide works best with very small test cases.
Landslide's coverage results from exploring the different scheduling possibilities that arise from a short sequence of system calls.
If Landslide were used when running a stress test, the resulting state space would be too large to explore feasibly.
Hence, when working with students, we provided a suite of four test programs, each no longer than 10 lines of code, but carefully written to bring about system call interactions known to be troublesome.
